% \usepackage{listings}
% Copyright 2020-2025 Robert Bosch GmbH

% Licensed under the Apache License, Version 2.0 (the "License");
% you may not use this file except in compliance with the License.
% You may obtain a copy of the License at

% http://www.apache.org/licenses/LICENSE-2.0

% Unless required by applicable law or agreed to in writing, software
% distributed under the License is distributed on an "AS IS" BASIS,
% WITHOUT WARRANTIES OR CONDITIONS OF ANY KIND, either express or implied.
% See the License for the specific language governing permissions and
% limitations under the License.

% --------------------------------------------------------------------------------------------------------------

\section{Overview}\label{sec:overview}
The \textbf{EventBusClient} provides a high-level API for interacting with a \textbf{RabbitMQ}-based event bus system.
It abstracts low-level AMQP details and supports dynamic loading of project-specific plugins for serializers, exchange handlers, and message structures.
This client is designed for modularity and ease of use across multiple projects.


\section{Features}\label{sec:features}
\begin{itemize}
  \item \textbf{Dynamic plugin loading}: Load custom serializers, exchange handlers, and message classes at runtime from a configurable plugins directory.
  \item \textbf{Thread-safe publishing}: Supports safe publishing from multiple threads with minimal locking overhead.
  \item \textbf{Automatic reconnect}: Automatically reconnects to \textbf{RabbitMQ} after connection loss (if enabled in configuration).
  \item \textbf{Pluggable architecture}: Supports both built-in and project-specific extensions.
  \item \textbf{Multiple exchange types}: Supports Topic, Fanout, Headers, and X-RTopic exchange handlers.
  \item \textbf{Header-based routing}: Route messages based on header attributes with AND/OR logic using HeadersExchangeHandler.
  \item \textbf{Supports JSON, Protobuf, Pickle serializers}.
\end{itemize}

\section{Architecture}\label{sec:architecture}
The \textbf{EventBusClient} is built around a modular architecture that allows for easy extension and customization. It consists of the following components:
\begin{itemize}
  \item \textbf{Base Classes}: Define interfaces for exchange handlers, serializers, and messages.
  \item \textbf{Plugins Directory}: Contains custom implementations of the base classes.
  \item \textbf{Configuration File}: Specifies the plugins path, \textbf{RabbitMQ} connection details, and selected implementations.
  \item \textbf{EventBusClient Class}: The main class that interacts with \textbf{RabbitMQ} and manages plugins.
  \item \textbf{Message Classes}: Define the structure of messages sent and received over the event bus.
  \item \textbf{Exchange Handlers}: Handle the declaration and publishing of messages to \textbf{RabbitMQ} exchanges.
  \item \textbf{Serializers}: Convert messages to and from byte streams for transmission over \textbf{RabbitMQ}.
\end{itemize}
The client uses a combination of built-in plugins and user-defined plugins to provide flexibility in message handling and serialization.
\section{Installation}\label{sec:installation}
To install the \textbf{EventBusClient}, use pip to install the package from PyPI:
\begin{pythonlog}
pip install eventbusclient
\end{pythonlog}
Ensure you have \textbf{RabbitMQ} server running and accessible at the specified host and port in the configuration file.
\section{Dependencies}\label{sec:dependencies}

The \textbf{EventBusClient} requires the following dependencies:
\begin{itemize}
  \item \textbf{pika}: For synchronous \textbf{RabbitMQ} communication
  \item \textbf{aio-pika}: Specifically for asynchronous \textbf{RabbitMQ} communication
  \item \textbf{protobuf}: For Protocol Buffers serialization (if using \textbf{ProtobufSerializer})
  \item \textbf{jsonpickle}: For JSON serialization (if using \textbf{JSONSerializer})
\end{itemize}
Ensure these dependencies are installed in your Python environment. You can install them using pip:
\begin{pythonlog}
pip install pika protobuf jsonpickle
\end{pythonlog}
\section{Directory Structure}\label{sec:directory-structure}
\begin{pythonlog}
project/
|-- EventBusClient/
|   |-- event_bus_client.py
|   |-- connection.py
|   |-- plugin_loader.py
|   |-- publisher.py
|   |-- subscriber.py
|   |-- qlogger.py
|   |-- message/
|   |   |-- base_message.py
|   |   |-- basic_message.py
|   |   |-- control_message.py
|   |-- exchange_handler/
|   |   |-- base.py
|   |   |-- topic_handler.py
|   |   |-- fanout_handler.py
|   |   |-- headers_handler.py
|   |   |-- x_rtopic_handler.py
|   |-- serializer/
|   |   |-- base_serializer.py
|   |   |-- json_serializer.py
|   |   |-- protobuf_serializer.py
|   |   |-- pickle_serializer.py
|   |-- plugins/
|   |   |-- custom_plugin/
|   |   |   |-- serializer/
|   |   |   |-- message/
|   |   |   |-- exchange_handler/
|   |-- config/
|   |   |-- config.jsonp
|-- app.py
\end{pythonlog}

\newpage
\section{Configuration}\label{sec:configuration}
The \plog{config.jsonp} file controls all aspects of the \textbf{EventBusClient}, including plugin paths, \textbf{RabbitMQ} connection, and selected implementations.

Configuration Template (\texttt{EventBusClient/config/config.jsonp.template})

\noindent
\textit{Note: This is JSON with comments (\texttt{.jsonp}). Remove comments (\texttt{//~, /*~*/}) if your parser does not support them.}

\begin{lstlisting}
{
  // plugins_path: Path to the directory containing plugin modules.
  "plugins_path": "./plugins",

  // host: Hostname or IP address for the server connection.
  "host": "localhost",

  // port: Port number for the server connection.
  "port": 5672,

  // serializer: Serialization method for message data (e.g., PickleSerializer).
  "serializer": "PickleSerializer",

  // exchange_handler: Handler type for message exchange (e.g., TopicExchangeHandler).
  "exchange_handler": "TopicExchangeHandler",

  // auto_reconnect: Automatically reconnect if the connection is lost (true/false).
  "auto_reconnect": true,

  // qos_prefetch: Number of messages to prefetch for Quality of Service.
  "qos_prefetch": 10,

  // general_cache_policy: "off" | "on_connect" | "on_demand".
  "general_cache_policy": "off",

  // general_routing_keys: Routing keys for general messages.
  "general_routing_keys": "general",

  // general_message_cls: Message class for general messages.
  "general_message_cls": "BaseMessage",

  // logger_name: Name of the logger instance.
  "logger_name": "test",

  // logfile: Path to the log file.
  "logfile": "test.log",

  // loglevel: Logging level for the logger instance.
  "loglevel": "INFO",

  // logger_mode: Log file mode ("w" for overwrite, "a" for append).
  "logger_mode": "a"
}
\end{lstlisting}

\vspace{1em}

\section*{How to Use}
\begin{enumerate}
  \item Copy the template above into your project as \texttt{config.jsonp}.
  \item Adjust each value to match your environment and requirements.
  \item Save the file and restart your application to apply changes.
\end{enumerate}

\section*{Field Descriptions}

\renewcommand{\arraystretch}{1.2}
\begin{longtable}{@{}p{4cm}p{2cm}p{2.2cm}p{3cm}p{6cm}@{}}
\toprule
\textbf{Field} & \textbf{Required} & \textbf{Type} & \textbf{Default} & \textbf{Description / Options} \\
\midrule
\endfirsthead
\toprule
\textbf{Field} & \textbf{Required} & \textbf{Type} & \textbf{Default} & \textbf{Description / Options} \\
\midrule
\endhead
\midrule
\multicolumn{5}{r}{\small Continued on next page}
\endfoot
\bottomrule
\endlastfoot

\texttt{plugins\_path} & Optional & String & \texttt{./plugins} & Path to the plugins directory (relative or absolute). \\

\texttt{host} & Optional & String & \texttt{"localhost"} & Server hostname or IP address for the broker connection. \\

\texttt{port} & Optional & Integer & \texttt{5672} & Server port for the broker connection. \\

\texttt{serializer} & \textbf{Mandatory} & String & (none) & Serialization strategy for message bodies (e.g., \texttt{PickleSerializer}). Must be available to both producer and consumer sides. \\

\texttt{exchange\_handler} & \textbf{Mandatory} & String & (none) & Exchange handler implementation (e.g., \texttt{TopicExchangeHandler}). Determines exchange type and binding logic. \\

\texttt{auto\_reconnect} & Optional & Boolean & \texttt{true} & Reconnect automatically if the connection is lost. \\

\texttt{qos\_prefetch} & Optional & Integer & \texttt{10} & Prefetch count for consumer QoS (how many unacked messages a consumer can receive). \\

\texttt{general\_cache\_policy} & Optional & String & \texttt{"off"} & Policy controlling the general message cache: \\
& & & &
\quad \texttt{"off"} -- disabled; \\
& & & &
\quad \texttt{"on\_connect"} -- start caching upon connection establishment; \\
& & & &
\quad \texttt{"on\_demand"} -- start caching only when first requested. \\

\texttt{general\_routing\_keys} & Optional & String (or list) & \texttt{"general"} & Routing key or keys used for the general cache subscription (supports topic patterns). \\

\texttt{general\_message\_cls} & Optional & String & \texttt{"BaseMessage"} & Message class to use for general-cache decoding. Provide a resolvable name or dotted path if needed. \\

\texttt{logger\_name} & Optional & String & \texttt{"event\_bus"} & Name of the logger instance to use/create. \\

\texttt{logfile} & Optional & String / Null & (none) & Path to a log file. If omitted or null, logs go to the default handlers (e.g., console). \\

\texttt{loglevel} & Optional & String & \texttt{"INFO"} & Logging threshold (\texttt{"DEBUG"}, \texttt{"INFO"}, \texttt{"WARNING"}, \texttt{"ERROR"}, \texttt{"CRITICAL"}). \\

\texttt{logger\_mode} & Optional & String & \texttt{"a"} & Log file mode: \texttt{"w"} (overwrite on each start) or \texttt{"a"} (append to existing). \\
\end{longtable}

\section{Usage}\label{sec:usage}

\subsection{Initialization}
Create an \textbf{EventBusClient} instance from configuration:
\begin{pythoncode}
from EventBusClient.event_bus_client import EventBusClient

client = await EventBusClient.from_config("./config/config.jsonp")
\end{pythoncode}


\subsection{Publishing Messages}
Send a message using the configured serializer and exchange handler asynchronously:
\begin{pythoncode}
msg = ListenerEventMsg(event="start_cycle", timestamp=1234567890)
await client.send("zone1.topic.start", msg, header)
\end{pythoncode}

Enable thread-safe publish from a non-async thread:
\begin{pythoncode}
client.send("zone1.topic.alert", msg, header, threadsafe=True)
\end{pythoncode}

Send a message synchronously using the corresponding method:
\begin{pythoncode}
client.send_sync("zone1.topic.sync", msg, header)
\end{pythoncode}


\subsection{Subscribing to Messages}
Subscribe to a routing key and handle incoming messages:
\begin{pythoncode}
async def on_message(msg: ListenerEventMsg, header):
    print(f"Received message: {msg}, header: {header}")

await client.on("zone1.#", ListenerEventMsg, on_message)
\end{pythoncode}


Subscribe to a routing key and handle incoming messages synchronously using \texttt{on\_sync}:
\begin{pythoncode}
def on_message_sync(msg: ListenerEventMsg, header):
    print(f"Received message: {msg}, header: {header}")

client.on_sync("zone1.#", ListenerEventMsg, on_message_sync)
\end{pythoncode}

\subsection{Header-Based Subscriptions (HeadersExchangeHandler)}
When using \textbf{HeadersExchangeHandler}, you can route messages based on header attributes instead of routing keys:

\begin{pythoncode}
# Subscribe with AND logic (all headers must match)
await client.on(
    routing_key="",  # Ignored in headers exchange
    message_cls=DocumentMessage,
    callback=on_document,
    binding_headers={"format": "pdf", "department": "engineering"},
    match_all=True  # x-match=all (AND logic)
)

# Subscribe with OR logic (any header must match)
await client.on(
    routing_key="",
    message_cls=DocumentMessage,
    callback=on_important,
    binding_headers={"format": "pdf", "priority": "high"},
    match_all=False  # x-match=any (OR logic)
)

# Publish with headers
await client.send(
    routing_key="",  # Ignored in headers exchange
    message=DocumentMessage(title="Q4 Report"),
    headers={"format": "pdf", "department": "engineering", "author": "john"}
)
\end{pythoncode}

\begin{itemize}
  \item \texttt{binding\_headers}: Dictionary of headers to match for subscription.
  \item \texttt{match\_all=True}: All specified headers must match (AND logic, x-match=all).
  \item \texttt{match\_all=False}: At least one header must match (OR logic, x-match=any).
\end{itemize}


\subsection{Setting Unroutable Message Policy}
You can configure how the client handles unroutable messages by setting a startup policy. For example:

\begin{pythoncode}
from EventBusClient.event_bus_client import EventBusClient
from EventBusClient.startup_policy import ConfigureUnroutablePolicy

config_unrout = ConfigureUnroutablePolicy(
    mode="return",                # "drop" (default), "alternate-exchange", or "return"
    alternate_exchange=None,      # Name of alternate exchange if mode is "alternate-exchange"
    on_unroutable="log",           # "log" (default), "raise", "cache", or "callback"
    on_unroutable_callback=None   # User-defined callback function if on_unroutable is "callback"
)
client = await EventBusClient.from_config(config_path, startup_policy=config_unrout)
\end{pythoncode}

\begin{itemize}
  \item \texttt{mode}: \emph{(optional, str, default: "drop")} \\
    Determines how unroutable messages are handled: \texttt{"drop"} (discard), \texttt{"alternate-exchange"} (send to alternate exchange), or \texttt{"return"} (return to sender).


  \item \texttt{alternate\_exchange}: \emph{(optional, str or None, default: None)} \\
    The name of the alternate exchange to use if mode is \texttt{"alternate-exchange"}.
  \item \texttt{on\_unroutable}: \emph{(optional, str, default: "log")} \\
    The action to take on unroutable messages: \texttt{"log"} (log the event), \texttt{"raise"} (raise an exception), \texttt{"cache"} (store for later), or \texttt{"callback"} (invoke a user callback).
  \item  \texttt{on\_unroutable\_callback}: \emph{(optional, callable or None, default: None)} \\
    A user-defined function to call when an unroutable message is received (only if \texttt{on\_unroutable} is set to \texttt{"callback"}).
\end{itemize}

\newpage
\section{Example: Producer and Consumer}
This example (from \plog{examples/basic_async_producer_consumer_sample.py}) starts a producer and a consumer in separate processes:
\begin{pythoncode}
# Start consumer
consumer = Process(target=run_process, args=(consumer_process, config_path))
consumer.start()

# Start producer
producer = Process(target=run_process, args=(producer_process, config_path))
producer.start()
# Wait for both to finish
producer.join()
consumer.join()
\end{pythoncode}

You can find the complete example in the \plog{examples} directory. Run run\_sample.bat (Windows) to execute the example.

\section{Example: Headers Exchange}
This example demonstrates header-based message routing using \textbf{HeadersExchangeHandler}:
\begin{pythoncode}
from EventBusClient.event_bus_client import EventBusClient
from EventBusClient.exchange_handler.headers_handler import HeadersExchangeHandler
from EventBusClient.serializer.json_serializer import JsonSerializer

# Create client with HeadersExchangeHandler
handler = HeadersExchangeHandler(
    name="documents_exchange",
    serializer=JsonSerializer()
)
client = EventBusClient(exchange_handler=handler, host="localhost", port=5672)
await client.connect()

# Subscribe to PDF documents from engineering (AND logic)
async def handle_docs(msg, headers):
    print(f"Received: {msg} with headers {headers}")

await client.on(
    routing_key="",
    message_cls=DocumentMessage,
    callback=handle_docs,
    binding_headers={"format": "pdf", "department": "engineering"},
    match_all=True
)

# Publish with matching headers
await client.send(
    routing_key="",
    message=DocumentMessage(title="Q4 Report"),
    headers={"format": "pdf", "department": "engineering"}
)
\end{pythoncode}

See \plog{examples/headers\_exchange\_sample.py} for the complete example.

\section{Example: Custom Plugin Structure}\label{sec:example:-custom-plugin-structure}
To create a custom plugin, follow these steps:
\begin{itemize}
  \item Create a directory for your plugin in the configured plugins path.
  \item Implement the required interfaces (e.g., BaseExchangeHandler, BaseMessage, BaseSerializer).
  \item Register your plugin in the \plog{config.jsonp} file.
  \item Ensure your plugin is discoverable by the \textbf{EventBusClient}.
  \item Place your plugin code in the plugins directory, following the structure:
  \item For example, if your plugin is named \plog{CustomPlugin}, the directory structure should look like this:
        \begin{pythonlog}
plugins/
|-- CustomPlugin/
    |-- __init__.py
    |-- custom_exchange_handler.py
    |-- custom_message.py
    |-- custom_serializer.py\end{pythonlog}
\end{itemize}
\section{Example: Built-in Plugins}
The \textbf{EventBusClient} comes with several built-in plugins that can be used directly or extended:

\subsection*{Exchange Handlers}
\begin{itemize}
  \item \textbf{TopicExchangeHandler}: Routes messages using routing key patterns with wildcards (* and \#).
  \item \textbf{FanoutExchangeHandler}: Broadcasts messages to all bound queues (ignores routing key).
  \item \textbf{HeadersExchangeHandler}: Routes messages based on header attributes with AND/OR logic.
  \item \textbf{XRTopicExchangeHandler}: Reverse topic exchange (subscribers use literal keys, publishers use patterns).
\end{itemize}

\subsection*{Message Classes}
\begin{itemize}
  \item \textbf{BasicMessage}: Simple message with content and headers.
  \item \textbf{ControlMessage}: Message for coordination and rendezvous patterns.
\end{itemize}

\subsection*{Serializers}
\begin{itemize}
  \item \textbf{PickleSerializer}: Serializes messages using Python's Pickle module.
  \item \textbf{JSONSerializer}: Serializes messages to JSON format.
  \item \textbf{ProtobufSerializer}: Serializes messages using Protocol Buffers.
\end{itemize}
These plugins can be used as-is or extended to create custom functionality.

\newpage
\section{Example: Built-in Configuration}
The built-in configuration file \plog{config.jsonp} provides a starting point for configuring the \textbf{EventBusClient}:
\begin{pythoncode}
{
  "plugins_path": "./plugins",
  "host": "localhost",
  "port": 5672,
  "serializer": "PickleSerializer",
  "exchange_handler": "XRTopicExchangeHandler",
  "message_class": "ListenerEventMsg",
  "threadsafe_publish": true,
  "auto_reconnect": true,
  "qos_prefetch": 10
}
\end{pythoncode}

This configuration specifies the plugins path, \textbf{RabbitMQ} connection details, serializer, exchange handler, message class, and other options.


\section{Example: Custom Exchange Handler}\label{sec:example:-custom-exchange-handler}
To create a custom exchange handler, implement the required interface and place it in the plugins directory. For example, a custom exchange handler might look like this:
\begin{pythoncode}
from event_bus.plugins import BaseExchangeHandler
class CustomExchangeHandler(BaseExchangeHandler):
    def declare_exchange(self, channel):
        # Custom exchange declaration logic
        pass

    def publish(self, channel, routing_key, message):
        # Custom publish logic
        pass
# Register the plugin in config.jsonp
{
  "plugins_path": "./plugins",
  "exchange_handler": "CustomExchangeHandler"
}
\end{pythoncode}

\section{Example: Custom Message Class}\label{sec:example:-custom-message-class}
To create a custom message class, define it in your project and register it in the configuration:
\begin{pythoncode}
from event_bus.plugins import BaseMessage
class CustomMessage(BaseMessage):
    def __init__(self, data):
        self.data = data

    @classmethod
    def from_data(cls, data):
        pass

    @abstractmethod
    def get_value(self):
        pass
# Register the message class in config.jsonp
{
  "plugins_path": "./plugins",
  "message_class": "CustomMessage"
}
\end{pythoncode}

\section{Example: Custom Serializer}\label{sec:example:-custom-serializer}
To create a custom serializer, implement the required interface and place it in the plugins directory. For example, a custom serializer might look like this:
\begin{pythoncode}
from event_bus.plugins import BaseSerializer
class CustomSerializer(BaseSerializer):
    def serialize(self, obj):
        # Custom serialization logic
        pass

    def deserialize(self, data):
        # Custom deserialization logic
        pass
# Register the plugin in config.jsonp
{
  "plugins_path": "./plugins",
  "serializer": "CustomSerializer"
}
\end{pythoncode}
